---
title: "Mountain Project Climbing Route Recommender"
author: "Mason Barnocky"
format:
  dashboard:
    theme: cosmo
    nav-buttons:
      - icon: github
        href: https://github.com/mojo1019/STAT408Final
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

library(dplyr)
library(stringr)
library(ggplot2)
library(leaflet)
library(DT)

US_STATES <- c(
  "Alabama","Alaska","Arizona","Arkansas","California","Colorado",
  "Connecticut","Delaware","Florida","Georgia","Hawaii","Idaho",
  "Illinois","Indiana","Iowa","Kansas","Kentucky","Louisiana","Maine",
  "Maryland","Massachusetts","Michigan","Minnesota","Mississippi",
  "Missouri","Montana","Nebraska","Nevada","New Hampshire","New Jersey",
  "New Mexico","New York","North Carolina","North Dakota","Ohio",
  "Oklahoma","Oregon","Pennsylvania","Rhode Island","South Carolina",
  "South Dakota","Tennessee","Texas","Utah","Vermont","Virginia",
  "Washington","West Virginia","Wisconsin","Wyoming"
)

TIER_COLS <- c(
  "Classic"  = "#E8593C",
  "Solid"    = "#EF9F27",
  "Good"     = "#3B8BD4",
  "Standard" = "#B4B2A9"
)

TYPE_COLS <- c(
  "Sport"   = "#185FA5",
  "Trad"    = "#1D9E75",
  "Boulder" = "#E8593C",
  "Alpine"  = "#7F77DD",
  "Other"   = "#B4B2A9"
)

parse_grade_numeric <- function(rating) {
  r        <- str_trim(rating)
  base_num <- suppressWarnings(as.integer(str_match(r, "5\\.(\\d+)")[, 2]))
  mod      <- str_extract(r, "[abcd+\\-]$")
  dplyr::case_when(
    is.na(base_num) ~ NA_real_,
    mod == "a" ~ base_num * 10 + 0,
    mod == "b" ~ base_num * 10 + 2,
    mod == "c" ~ base_num * 10 + 5,
    mod == "d" ~ base_num * 10 + 8,
    mod == "+" ~ base_num * 10 + 9,
    mod == "-" ~ base_num * 10 - 1,
    TRUE       ~ base_num * 10 + 4
  )
}

primary_type <- function(type_str) {
  dplyr::case_when(
    str_detect(type_str, regex("boulder", ignore_case = TRUE)) ~ "Boulder",
    str_detect(type_str, regex("sport",   ignore_case = TRUE)) ~ "Sport",
    str_detect(type_str, regex("trad",    ignore_case = TRUE)) ~ "Trad",
    str_detect(type_str, regex("alpine",  ignore_case = TRUE)) ~ "Alpine",
    TRUE ~ "Other"
  )
}

clean_col_names <- function(df) {
  names(df) <- trimws(names(df))
  df <- df[, !(names(df) == "" | grepl("^Unnamed", names(df))), drop = FALSE]
  names(df)[names(df) == "Avg Stars"]      <- "avg_stars"
  names(df)[names(df) == "Route Type"]     <- "route_type"
  names(df)[names(df) == "Pitches"]        <- "pitches"
  names(df)[names(df) == "Area Latitude"]  <- "lat"
  names(df)[names(df) == "Area Longitude"] <- "lon"
  names(df)[names(df) == "Rating"]         <- "rating"
  names(df)[names(df) == "num_votes"]      <- "num_votes"
  df
}

build_routes <- function(raw) {
  raw        <- clean_col_names(raw)
  parts_list <- str_split(raw$Location, " > ")
  raw$state  <- vapply(parts_list, function(p) { p <- trimws(p); p[length(p)] }, character(1))
  raw$crag   <- vapply(parts_list, function(p) trimws(p)[1], character(1))
  raw |>
    filter(state %in% US_STATES,
           !is.na(lat), !is.na(lon),
           lat >= 18, lat <= 72, lon >= -180, lon <= -66) |>
    mutate(
      grade_num  = parse_grade_numeric(rating),
      climb_type = primary_type(route_type),
      num_votes  = pmax(1L, suppressWarnings(as.integer(num_votes)), na.rm = TRUE),
      avg_stars  = as.numeric(avg_stars),
      pitches    = pmax(1L, suppressWarnings(as.integer(pitches)), na.rm = TRUE),
      is_classic = avg_stars >= 3.8 & num_votes >= 20,
      route_tier = dplyr::case_when(
        avg_stars >= 3.8 & num_votes >= 20 ~ "Classic",
        avg_stars >= 3.2 & num_votes >= 10 ~ "Solid",
        avg_stars >= 2.5                   ~ "Good",
        TRUE                               ~ "Standard"
      ),
      grade_band = cut(grade_num,
        breaks = c(-Inf, 55, 79, 85, 99, 105, 109, 115, 119, 124, 129, Inf),
        labels = c("5.5","5.6-5.7","5.8","5.9","5.10a/b","5.10c/d",
                   "5.11a/b","5.11c/d","5.12a/b","5.12c/d","5.13+"))
    )
}

score_routes <- function(df, grade_mid) {
  C_global <- mean(df$avg_stars, na.rm = TRUE)
  df |>
    mutate(
      bayesian_q    = (num_votes / (num_votes + 10)) * avg_stars +
                      (10 / (num_votes + 10)) * C_global,
      grade_match   = exp(-0.5 * ((grade_num - grade_mid) / 15)^2),
      pitch_bonus   = dplyr::case_when(
        pitches > 5 ~ 0.25, pitches > 1 ~ 0.10, TRUE ~ 0.00
      ),
      classic_bonus = ifelse(is_classic, 0.20, 0),
      score = round(
        (bayesian_q * 0.45 + grade_match * 0.30 +
         pitch_bonus * 0.10 + classic_bonus * 0.15) * (10 / 4), 1
      )
    ) |>
    arrange(desc(score))
}

# Load data — prefer pre-built RDS, fall back to CSV
rds_path <- "mp_routes.rds"
csv_path <- "mp_routes.csv"

if (file.exists(rds_path)) {
  routes <- readRDS(rds_path)
  if (any(is.na(names(routes)) | names(routes) == "")) {
    routes <- build_routes(read.csv(csv_path, stringsAsFactors = FALSE,
                                    check.names = FALSE))
    saveRDS(routes, rds_path)
  }
} else if (file.exists(csv_path)) {
  routes <- build_routes(read.csv(csv_path, stringsAsFactors = FALSE,
                                  check.names = FALSE))
  saveRDS(routes, rds_path)
} else {
  stop("Place mp_routes.rds or mp_routes.csv in the same folder as this .qmd file.")
}

GRADE_MIN <- min(routes$grade_num, na.rm = TRUE)
GRADE_MAX <- max(routes$grade_num, na.rm = TRUE)

# ── Pre-compute everything at render time ─────────────────

# Crag summary for map (top 500 by route count)
crag_map <- routes |>
  group_by(crag, state, lat, lon) |>
  summarise(
    route_count = n(),
    avg_stars   = round(mean(avg_stars, na.rm = TRUE), 2),
    n_classic   = sum(is_classic, na.rm = TRUE),
    top_grade   = rating[which.max(grade_num)],
    .groups     = "drop"
  ) |>
  filter(!is.na(lat), !is.na(lon)) |>
  arrange(desc(route_count)) |>
  slice_head(n = 500)

# Build route-list popups
visible_crags  <- crag_map$crag
routes_visible <- routes[routes$crag %in% visible_crags, ]
crag_split     <- split(routes_visible, routes_visible$crag)

build_popup <- function(i) {
  cn  <- crag_map$crag[i]
  cr  <- crag_split[[cn]]
  if (is.null(cr) || nrow(cr) == 0) return("")
  C  <- mean(cr$avg_stars, na.rm = TRUE)
  cr <- cr |>
    mutate(bq = (num_votes / (num_votes + 10)) * avg_stars +
                (10 / (num_votes + 10)) * C) |>
    arrange(desc(bq))
  top8 <- head(cr, 8)
  tier_color <- function(t) switch(t,
    "Classic" = "#E8593C", "Solid" = "#EF9F27",
    "Good" = "#3B8BD4", "#B4B2A9"
  )
  rows <- paste(vapply(seq_len(nrow(top8)), function(j) {
    r  <- top8[j, ]
    tc <- tier_color(r$route_tier)
    nm <- if (!is.na(r$URL) && nchar(r$URL) > 0)
      sprintf('<a href="%s" target="_blank" style="color:#185FA5;font-weight:500;text-decoration:none">%s</a>', r$URL, r$Route)
    else sprintf('<span style="font-weight:500">%s</span>', r$Route)
    sprintf(
      '<div style="padding:4px 0;border-bottom:1px solid #f0ede8">
         %s<span style="float:right;font-size:10px;background:%s;color:#fff;padding:1px 5px;border-radius:3px">%s</span><br>
         <span style="font-size:11px;color:#777">%s | %s | %.1f stars</span>
       </div>',
      nm, tc, r$route_tier, r$rating, r$climb_type, r$avg_stars
    )
  }, character(1)), collapse = "")
  more <- if (crag_map$route_count[i] > 8)
    sprintf('<div style="margin-top:6px;font-size:11px;color:#888;text-align:center">+ %d more routes on Mountain Project</div>',
            crag_map$route_count[i] - 8)
  else ""
  sprintf(
    '<div style="font-family:sans-serif;font-size:13px;min-width:240px;max-width:300px">
       <b style="font-size:15px">%s</b><br>
       <span style="color:#888;font-size:12px">%s</span>
       <hr style="margin:6px 0">
       <div style="font-size:12px;color:#555;margin-bottom:8px">
         <b>%d</b> route%s | avg <b>%.1f</b> stars | hardest <b>%s</b>
       </div>%s%s</div>',
    cn, crag_map$state[i],
    crag_map$route_count[i], ifelse(crag_map$route_count[i] == 1, "", "s"),
    crag_map$avg_stars[i], crag_map$top_grade[i], rows, more
  )
}

popup_html <- vapply(seq_len(nrow(crag_map)), build_popup, character(1))

# Scores for recommendations page (intermediate default ~5.10)
rec_scored <- score_routes(routes, grade_mid = 99)

rec_top10 <- rec_scored |>
  slice_head(n = 10) |>
  mutate(
    name_short = str_trunc(Route, 28, ellipsis = "..."),
    wall_short = str_trunc(crag,  22, ellipsis = "..."),
    bar_label  = paste0(rating, " \u00b7 ", wall_short),
    label      = factor(name_short, levels = rev(name_short))
  )

rec_top100 <- rec_scored |>
  slice_head(n = 100) |>
  mutate(
    Route = ifelse(
      !is.na(URL) & nchar(URL) > 0,
      paste0('<a href="', URL, '" target="_blank">', Route, '</a>'),
      Route
    ),
    Stars = round(avg_stars, 1)
  ) |>
  select(Route, Wall = crag, State = state, Grade = rating,
         Stars, Votes = num_votes, Type = climb_type,
         Pitches = pitches, Score = score, Tier = route_tier)
```

# About the Data

## Row {height="20%"}

```{r}
#| content: valuebox
#| title: "US Routes"
list(icon = "map", color = "#185FA5",
     value = format(nrow(routes), big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "Climbing Walls"
list(icon = "geo-alt", color = "#1D9E75",
     value = format(n_distinct(routes$crag), big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "States"
list(icon = "flag", color = "#7F77DD",
     value = n_distinct(routes$state))
```

```{r}
#| content: valuebox
#| title: "Classic Routes"
list(icon = "star", color = "#EF9F27",
     value = format(sum(routes$is_classic), big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "Avg Star Rating"
list(icon = "bar-chart", color = "#E8593C",
     value = round(mean(routes$avg_stars, na.rm = TRUE), 1))
```

## Row {height="80%"}

### Column {width="55%"}

```{r}
#| title: "Where are the climbs? (top 500 walls)"
#| padding: 0

pal <- colorNumeric(
  palette = c("#3B8BD4","#1D9E75","#EF9F27","#E8593C"),
  domain  = c(0, 4)
)

leaflet(crag_map, options = leafletOptions(minZoom = 3)) |>
  addProviderTiles("CartoDB.Positron") |>
  setView(lng = -98, lat = 39, zoom = 4) |>
  addCircleMarkers(
    lng = ~lon, lat = ~lat,
    radius      = ~pmax(4, pmin(sqrt(route_count) * 2, 20)),
    fillColor   = ~pal(avg_stars),
    fillOpacity = 0.75,
    color       = "#ffffff",
    weight      = 1.2,
    popup       = popup_html,
    label       = ~paste0(crag, " (", route_count, " routes)"),
    clusterOptions = markerClusterOptions(
      maxClusterRadius = 40, showCoverageOnHover = FALSE
    )
  ) |>
  addLegend("bottomright",
    colors  = c("#3B8BD4","#1D9E75","#EF9F27","#E8593C"),
    labels  = c("0-1","1-2","2-3","3-4"),
    title   = "Avg stars", opacity = 0.9
  )
```

### Column {width="45%"}

::: {.card title="About this project"}
This dashboard explores nearly **90,000 climbing routes** across the United States sourced from [Mountain Project](https://www.mountainproject.com), the largest crowd-sourced climbing database in North America.

The goal is to help climbers answer: *Where should I climb? What routes match my ability? What are the best climbs near me?*

**What the data contains:**

- Route name, grade (YDS), and type — sport, trad, boulder, alpine
- Average star rating and number of votes from the climbing community
- GPS coordinates, pitch count, and route length
- Location hierarchy: wall > sub-area > region > state

**What this dashboard covers:**

- **About the Data** — dataset overview and geographic distribution
- **Analysis** — grade distributions and quality patterns
- **Recommendations** — top-scored routes using the Bayesian quality model
- **Summary** — key findings and links to further resources

For a fully interactive version with real-time filters by grade, state, and style, see the companion **[Shiny app](https://github.com/mojo1019/STAT408Final)** in the GitHub repo.
:::

```{r}
#| title: "Routes by state (top 15)"

routes |>
  count(state, sort = TRUE) |>
  slice_head(n = 15) |>
  mutate(state = factor(state, levels = rev(state))) |>
  ggplot(aes(x = state, y = n)) +
  geom_col(fill = "#185FA5", width = 0.7) +
  geom_text(aes(label = format(n, big.mark = ",")),
            hjust = -0.1, size = 3, color = "#3d3d3a") +
  coord_flip(clip = "off") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = NULL, y = "Number of routes") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank())
```

# Analysis

## Row

### Column {width="50%"}

```{r}
#| title: "Grade distribution by quality tier"

routes |>
  filter(!is.na(grade_band)) |>
  count(grade_band, route_tier) |>
  mutate(route_tier = factor(route_tier, levels = names(TIER_COLS))) |>
  ggplot(aes(x = grade_band, y = n, fill = route_tier)) +
  geom_col(width = 0.75) +
  coord_flip() +
  scale_fill_manual(values = TIER_COLS, name = "Quality tier") +
  labs(x = NULL, y = "Number of routes",
       caption = "Classic = 3.8+ stars with 20+ votes") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank(),
        legend.position    = "bottom")
```

### Column {width="50%"}

```{r}
#| title: "Route type breakdown by quality"

routes |>
  count(climb_type, route_tier) |>
  filter(!is.na(climb_type)) |>
  mutate(route_tier = factor(route_tier, levels = names(TIER_COLS))) |>
  ggplot(aes(x = reorder(climb_type, n, sum), y = n, fill = route_tier)) +
  geom_col(width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = TIER_COLS, name = "Quality tier") +
  labs(x = NULL, y = "Number of routes") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank(),
        legend.position    = "bottom")
```

## Row

### Column {width="50%"}

```{r}
#| title: "Star rating distribution"

mean_stars <- round(mean(routes$avg_stars, na.rm = TRUE), 2)
med_stars  <- round(median(routes$avg_stars, na.rm = TRUE), 2)

routes |>
  filter(!is.na(avg_stars)) |>
  ggplot(aes(x = avg_stars)) +
  geom_histogram(binwidth = 0.25, fill = "#EF9F27", color = "white",
                 linewidth = 0.3) +
  geom_vline(xintercept = mean_stars, linetype = "dashed",
             color = "#E8593C", linewidth = 0.9) +
  annotate("text", x = mean_stars + 0.1, y = Inf, vjust = 2,
           hjust = 0, size = 3.2, color = "#E8593C",
           label = paste0("mean = ", mean_stars)) +
  labs(x = "Average star rating", y = "Number of routes",
       caption = paste0("Median = ", med_stars)) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

### Column {width="50%"}

```{r}
#| title: "Summary statistics by climbing type"

routes |>
  group_by(Type = climb_type) |>
  summarise(
    Routes       = n(),
    `Avg Stars`  = round(mean(avg_stars, na.rm = TRUE), 2),
    Classics     = sum(is_classic),
    `% Classic`  = paste0(round(mean(is_classic) * 100, 1), "%"),
    `Avg Pitches`= round(mean(pitches, na.rm = TRUE), 1),
    .groups = "drop"
  ) |>
  arrange(desc(Routes)) |>
  datatable(
    rownames = FALSE,
    options  = list(dom = "t", pageLength = 10,
                    order = list(list(1L, "desc")))
  ) |>
  formatStyle("Avg Stars",
    background         = styleColorBar(c(0, 4), "#FAEEDA"),
    backgroundSize     = "100% 80%",
    backgroundRepeat   = "no-repeat",
    backgroundPosition = "center"
  )
```

# Recommendations

## Row

### Column {width="40%"}

```{r}
#| title: "Top 10 recommended climbs"
#| fig-height: 6

ggplot(rec_top10, aes(x = label, y = score, fill = climb_type)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = paste0(score, "/10")),
            hjust = -0.1, size = 3.0, color = "#3d3d3a") +
  geom_text(aes(label = bar_label, y = 0.25),
            hjust = 0, size = 2.8, color = "white") +
  coord_flip(clip = "off") +
  scale_fill_manual(values = TYPE_COLS, name = "Type") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.22))) +
  labs(x = NULL, y = "Score (0-10)",
       caption = "Default: intermediate grade range (5.9-5.10). Full filtering available in the Shiny app.") +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    legend.position    = "bottom",
    axis.text.y        = element_text(size = 10)
  )
```

::: {.card title="How routes are scored"}
Routes are ranked using a composite score (0-10) combining four signals:

- **Bayesian quality (45%)** — star rating adjusted for vote count; low-vote routes are pulled toward the global average so a single 4-star vote doesn't outrank a 3.9-star route with 500 votes
- **Grade match (30%)** — peaks at the target experience level and falls off smoothly for routes that are too easy or too hard
- **Classic bonus (15%)** — extra credit for routes with 3.8+ stars and 20+ community votes
- **Pitch bonus (10%)** — small boost for multi-pitch routes

*For live filtering by grade, state, and style, use the companion [Shiny app](https://github.com/mojo1019/STAT408Final).*
:::

### Column {width="60%"}

```{r}
#| title: "Top 100 routes by recommendation score"

datatable(rec_top100,
  escape   = FALSE,
  rownames = FALSE,
  options  = list(
    pageLength = 15,
    dom        = "ftp",
    order      = list(list(8L, "desc")),
    columnDefs = list(
      list(className = "dt-center", targets = c(3, 4, 5, 6, 7, 8))
    )
  )
) |>
  formatStyle("Score",
    background         = styleColorBar(c(0, 10), "#B5D4F4"),
    backgroundSize     = "100% 80%",
    backgroundRepeat   = "no-repeat",
    backgroundPosition = "center"
  ) |>
  formatStyle("Tier",
    backgroundColor = styleEqual(
      c("Classic","Solid","Good","Standard"),
      c("#FAEEDA","#FFF5E0","#E6F1FB","#F1EFE8")
    ),
    color = styleEqual(
      c("Classic","Solid","Good","Standard"),
      c("#633806","#4A2E00","#042C53","#444441")
    )
  )
```

# Summary

## Row

### Column {width="60%"}

::: {.card title="What this project found"}
This dashboard set out to answer a simple question that every climber faces: **where should I climb, and what routes are right for me?** After exploring nearly 90,000 routes across the United States, several clear patterns emerged.

**The data is heavily skewed toward the West.** Colorado, California, and Utah together account for over a third of all routes. This reflects both the concentration of quality climbing terrain and the demographics of Mountain Project's user base.

**Moderate grades dominate.** The 5.9-5.10 range has by far the most routes, with Classic-tier quality concentrated in the 5.10-5.11 window. The sweet spot for finding high-quality climbing with lots of options is the intermediate grade range.

**Star ratings are unreliable without vote context.** Many routes have high average stars but only one or two votes. The Bayesian scoring approach addresses this by pulling low-vote ratings toward the global mean of ~3.2 stars, so popular well-reviewed routes rank above obscure ones with a single lucky rating.

**Sport climbing has the most routes; trad has the highest classic rate.** Despite sport climbing dominating in raw numbers, trad routes punch above their weight in the Classic tier — reflecting the longer history and stronger community curation around trad destinations.

**The recommendation system meaningfully reorders results.** Simple star-rating filters return routes that look good on paper but may be at the wrong grade or poorly documented. The composite score — combining quality, grade fit, pitch count, and classic status — surfaces routes that are both high quality and appropriate for the user's level. The full interactive version of this system is available in the companion Shiny app.
:::

### Column {width="40%"}

::: {.card title="Further resources"}
**Climbing data & community**

- [Mountain Project](https://www.mountainproject.com) — source dataset; route pages, area guides, conditions reports
- [The Crag](https://www.thecrag.com) — alternative global climbing database
- [OpenBeta](https://openbeta.io) — open-source climbing database with a public API

**Methods**

- [Bayesian average ratings](https://www.evanmiller.org/bayesian-average-ratings.html) — the theory behind the weighted quality score used in this project
- [Leaflet for R](https://rstudio.github.io/leaflet/) — interactive map documentation
- [Quarto dashboards](https://quarto.org/docs/dashboards/) — how this dashboard was built

**R packages used**

- [`dplyr`](https://dplyr.tidyverse.org) + [`stringr`](https://stringr.tidyverse.org) — data wrangling and location parsing
- [`ggplot2`](https://ggplot2.tidyverse.org) — grade pyramid, type breakdown, star distribution
- [`leaflet`](https://rstudio.github.io/leaflet/) — interactive climbing wall map
- [`DT`](https://rstudio.github.io/DT/) — searchable route recommendation table

**GitHub repository**

Full source code, data, and companion Shiny app:
[github.com/mojo1019/STAT408Final](https://github.com/mojo1019/STAT408Final)
:::
